### 2023-09-25 chapters/discussion.tex
Work on background and main contributions

\section{\Cref{chp:paper-review} revisited: \dots}

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}

@@

>>>>

\section{\Cref{chp:paper-review} revisited: \dots} \label{sec:discussion-paper-review}

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation? \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}

@@

\lesstodo[inline]{Maybe mention MultiQT paper \cite{havtorn_multiqt_2020}.}

### 2023-09-25 chapters/main-contributions.tex
First complete introduction
Work on background and main contributions

The chapters of the thesis are self-contained studies and therefore detail their own contributions. 
They are however written without consideration to the other chapters. 
For that reason, this chapter will provide a high-level overview of the organization of the thesis and describe its main
contributions in relation to the overall research topic.

\contribitem{part:background}
This is the part of the thesis you are currently reading. \Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics. \Cref{chp:technical-background} provided the necessary technical background for the rest of the thesis by introducing uncertainty as a concept in the context of information theory and probability theory and providing a brief overview of generative latent variable models. 
\todo[inline]{Update the contributions of the technical background section accordingly.}

\contribitem{part:unsupervised-uncertainty-estimation}
This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. The first two papers are concerned with out-of-distribution detection using generative models. The first paper proposes a new method for out-of-distribution detection using hierarchical latent variable models. The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. The third paper proposes a new hierarchical latent variable model for speech

\contribitem{chp:paper-hierarchical}
This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical latent variable models. The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions. To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. The proposed method is computationally efficient, fully unsupervised, and performs well on several image-based out-of-distribution detection benchmarks.
Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.

\contribitem{chp:paper-modelagnostic}
This paper builds on the previous paper and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. The method is based on combining a classical parametric test with the recently introduced typicality test using Fisher's method. We show that this leads to a more accurate out-of-distribution test. We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical out-of-distribution detection, especially in high-risk applications such as in medical devices or services.

\contribitem{chp:paper-benchmarking}
This paper proposes a new hierarchical latent variable model for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. The model is benchmarked against other latent variable models and autoregressive models for speech abnd we show that using a hierarchy of latent variables improves the likelihood of the model. Finally, we show that the learned latent space captures information relevant for speech recognition such as phonetic content.

\contribitem{part:self-supervised-speech-representation-learning}
This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \cite{mohamed_selfsupervised_2022}. 
Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic methods such as those investigated in \cref{part:unsupervised-uncertainty-estimation}. 
In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation. We shall return to this in the discussion of \cref{chp:paper-review}.

\todo[inline]{Update this paragraph according to the discussion of self-supervised learning for uncertainty estimation}
% \cite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.

@@

\todo[inline]{Should we include the brief review paper as a chapter instead of the long review paper?}

@@

Prior to this thesis, we published related work on question tracking for calls to emergency services using a multimodal approach that combines automatic transcripts with the raw audio \cite{havtorn_multiqt_2020}.

\contribitem{chp:paper-retrospective}
Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls. 
We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline. First, calls are transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.

\contribitem{chp:paper-automated}

>>>>

The following chapters of the thesis are self-contained studies and therefore detail their own contributions. 
They are however written without consideration to the other chapters. 
For that reason, this chapter will describe the main contributions of each chapter in relation to the overall research topic of the thesis. 
The chapters are structured into three parts: \cref{part:unsupervised-uncertainty-estimation} is concerned with unsupervised uncertainty estimation, \cref{part:self-supervised-speech-representation-learning} is concerned with self-supervised speech representation learning, and \cref{part:medical-applications} contains studies on machine learning methods applied for tasks in a medical setting. 
Finally, \cref{part:discussion-and-conclusion} contains the discussion and conclusions of the thesis.

\contribitem{part:background}
\Cref{chp:introduction} provided a general introduction to the thesis and gave motivating examples for speech recognition and assistance in medical diagnostics. \Cref{chp:technical-background} provided additional technical background not included in the individual studies by introducing uncertainty as a concept in the context of information and probability theory, introducing the task of out-of-distribution detection and reviewing existing work on the problem, and providing an introduction to variational autoencoders.

\contribitem{part:unsupervised-uncertainty-estimation}
This part of the thesis is concerned with unsupervised uncertainty estimation and consists of three papers. 
The first two papers are concerned with out-of-distribution detection using generative models. The first paper proposes a new method for out-of-distribution detection using hierarchical variational autoencoders. 
The second paper builds on the first and proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable generative model with explicit likelihood. 
The third paper proposes a new hierarchical variational autoencoder for speech and benchmarks it against other variational autoencoders and autoregressive models.

\contribitem{chp:paper-hierarchical}
This paper proposes a new method for unsupervised out-of-distribution detection using hierarchical variational autoencoders. 
The paper provides evidence that the failure of deep generative models to assign lower likelihoods to data from outside the training distribution is due to overemphasis on low-level features that generalize between distributions. 
To solve this, the proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
% Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.

\contribitem{chp:paper-modelagnostic}
This paper proposes a model-agnostic method for out-of-distribution detection that can be used with any differentiable, explicit likelihood generative model. 
The method is based on combining a classical parametric test with the recently introduced typicality test via Fisher's method. 
We show that this leads to a more accurate out-of-distribution test. We also discuss the benefits of casting out-of-distribution detection as a statistical testing problem, for instance enabling false positive rate control which is valuable for practical applications, especially in high-risk settings such as medical decision-making.

\contribitem{chp:paper-benchmarking}
This paper proposes a new hierarchical variational autoencoder for speech inspired by the Clockwork VAE \cite{saxena_clockwork_2021}. 
The model is benchmarked against other variational autoencoders and autoregressive models for speech. We show that using a hierarchy of latent variables improves the likelihood of the model and that the learned latent space captures information relevant for speech recognition such as phonetic content.

\contribitem{part:self-supervised-speech-representation-learning}
This part of the thesis is concerned with self-supervised speech representation learning and consists of a single paper that provides a substantial review of the field \cite{mohamed_selfsupervised_2022}. 
Self-supervised learning is a promising approach to speech representation learning that has shown a fast, wide and successful adoption across speech modelling over the course of the last few years. 
As measured by performance on common downstream tasks such as speech recognition and spoken language understanding, self-supervised representations generally outperform representations learned via probabilistic generative models such as those investigated in \cref{part:unsupervised-uncertainty-estimation}. 
In this review paper, we do not consider uncertainty estimation, but in future work, consideration should be given to the use of self-supervised learning as an alternative paradigm to generative models for learning representations for unsupervised uncertainty estimation. We shall return to this in the discussion of \cref{chp:paper-review} in \cref{sec:discussion-paper-review}.

@@

@@

\contribitem{chp:paper-retrospective}
Effective stroke treatment depends on recognition by call-takers at prehospital telehealth services. This paper aimed to develop and assess the potential of machine learning in improving prehospital stroke recognition during medical helpline calls. 
We used calls from Copenhagen during 2015 to 2020, to develop a machine learning-based classification pipeline. First, calls were transcribed by a speech recognition model and then categorised as stroke or non-stroke using a text classification model.
On test data from 2021, call-takers achieved an overall sensitivity of 52.7\% (95\% confidence interval 49.2-56.4\%) with a positive predictive value (PPV) of 17.1\% (15.5-18.6\%) while the machine learning framework performed significantly better (p < 0.0001) with a sensitivity of 63.0\% (62.0-64.1\%) and a PPV of 24.9\% (24.3-25.5\%).
This study provides preliminary evidence that a machine learning framework for recognizing stroke in prehospital medical helpline calls could become a supportive tool for call-takers, aiding in early and accurate stroke recognition.

\contribitem{chp:paper-automated}
Improving the efficiency of medical coding can alleviate the administrative burden on medical staff and help reduce the rising costs for hospitals. 
In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is often attributable to weak configuration, crudely sampled train-test splits, and insufficient evaluation. 
We compare models on data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
By analyzing prediction errors, we confirm that all models struggle with rare codes, while long documents only have a negligible impact. 
We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models. 
This paper aims to increase reproducibility and comparability of future work in automated medical coding and to provide a baseline for future research.

### 2023-09-25 chapters/paper_hierarchical.tex
Work on background and main contributions

Since the scores in \cref{eq_hierarchical:llr-as-kls,eq_hierarchical:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
Since $\log p_\thetab(\xb)$ and part of the KL divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance. 
Empirical results indeed show that $\text{Var}(\widehat{LLR}^{>k})$ is larger than $\text{Var}(\hat{\mathcal{L}})$ but smaller than $\text{Var}(\hat{\mathcal{L}}^{>k})$.
%Whether this decreases the variance to below the variance of any of the two bounds, we do not know, and we believe it to be difficult to verify mathematically.
Nevertheless, the variance of the estimators is guaranteed to go to zero as the number of samples is increased.

>>>>

Since the scores in \cref{eq_hierarchical:llr-as-kls,eq_hierarchical:llr-as-kls-iwae-reduced} are estimated by sampling their estimators are stochastic objects with nonzero variance.
We note that $\text{Var}(\widehat{LLR}^{>k}) = \text{Var}(\hat{\mathcal{L}}) + \text{Var}(\hat{\mathcal{L}}^{>k}) - 2\, \text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$.
Since $\log p_\thetab(\xb)$ and part of the KL-divergence are identical in the expressions of $\mathcal{L}$ and $\mathcal{L}^{>k}$ we expect $\text{Cov}(\hat{\mathcal{L}}, \hat{\mathcal{L}}^{>k})$ to be positive which reduces the total variance. 
Empirical results indeed show that $\text{Var}(\widehat{LLR}^{>k})$ is larger than $\text{Var}(\hat{\mathcal{L}})$ but smaller than $\text{Var}(\hat{\mathcal{L}}^{>k})$.
%Whether this decreases the variance to below the variance of any of the two bounds, we do not know, and we believe it to be difficult to verify mathematically.
Nevertheless, the variance of the estimators is guaranteed to go to zero as the number of samples is increased.

### 2023-09-25 chapters/paper_review/02_three_waves.tex
Add and revise short table and figure captions
Work on background and main contributions

\section{Historical Context of Representation Learning}

@@

More recently, representation learning has seen a shift of focus towards neural models,
which, compared to GMMs and HMMs, offer distributed representations with more
capacity to model diverse input signals into efficient latent binary codes. 
Examples of early techniques include restricted Boltzmann machines
(RBM)~\cite{hinton_reducing_2006}, denoising autoencoders~\cite{vincent_stacked_2010}, noise contrastive
estimation (NCE)~\cite{gutmann_noisecontrastive_2012}, sparse coding~\cite{olshausen_emergence_1996,
lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\cite{ranzato_unified_2007}.
Many of these techniques have also been applied to CV and NLP problems, which
provided inspiration for their application to speech.

>>>>

\section{Historical context of representation learning}

@@

More recently, representation learning has seen a shift of focus towards neural models,
which, compared to GMMs and HMMs, offer distributed representations with more
capacity to model diverse input signals into efficient latent binary codes. 
Examples of early techniques include restricted Boltzmann machines
(RBM)~\cite{hinton_reducing_2006}, denoising autoencoders~\cite{vincent_stacked_2010}, noise contrastive
estimation (NCE)~\cite{gutmann_noisecontrastive_2010}, sparse coding~\cite{olshausen_emergence_1996,
lee_efficient_2006, sivaram_sparse_2010}, and energy-based methods~\cite{ranzato_unified_2007}.
Many of these techniques have also been applied to CV and NLP problems, which
provided inspiration for their application to speech.

### 2023-09-25 chapters/paper_review/03_approaches.tex
Work on background and main contributions

\subsubsection{Approaches}
\paragraph{CPC}
\label{par:cpc}
Contrastive Predictive Coding (CPC)~\cite{oord_representation_2018} is a prominent example of a contrastive model. 
\edit{CPC uses a convolutional module $f_1(\cdot)$ to produce localized representations $z_t$ with a recurrent module $f_2(\cdot)$ on top that outputs a contextualized representation $h_t$. An anchor representation $\hat{z}_{t,k}$ is obtained via a linear projection $g_k(\cdot)$ of $h_t$. The positives and negatives are sampled from the localized representation $Z$.
Hence, at a single timestep $t$, CPC forms multiple anchor representations $\hat{z}_{t,k}$ for $k\in\{1,\dots,K\}$ and associates with each one a single positive sample at the corresponding timestep, $z_{t+k}$, $k$ steps in the future:
\begin{align}
    z_t &= f_{1}(X_{[t-u,t+u]}) \label{eq: cpc local representation} \enspace ,\\
    H_{[1,t]} &= f_{2}(Z_{[1,t]}) \enspace , \\
    \hat{z}_{t,k} &= g_k(h_{t}) \enspace.
\end{align}
%
\noindent Each $z_{t}$ only encodes information from a limited receptive field, while $f_2(\cdot)$ is limited to condition each $h_t$ on previous timesteps $Z_{[1,t]}$. \ci{Without these restrictions}, the model could collapse to a trivial solution. $g_k$ is a unique transformation per \ci{offset} $k$ (e.g., a linear projection).
The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2012}. \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:
\begin{align}
    \mathcal{L}_{t,k} &= - \log \left(\frac{\exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})}{\sum_{\ci{i \in \mathcal{I}}} \exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})} \right)\enspace .
    \label{eq: cpc loss}
\end{align}
\noindent \edit{Here, $\mathcal{I}$ \ci{is a random subset of $N$} indices  which includes the target index $t+k$ and $N-1$ negative samples drawn from a proposal distribution, e.g., a uniform distribution over $\{1,\dots,T\}$. Including the target index in $\mathcal{I}$ ensures that the loss is a proper categorical cross-entropy and that minimizing it has the previously stated relation to mutual information maximization}. This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord_representation_2018}. The loss is indexed by $k$ to show that CPC targets multiple offsets using different projection layers $g_k(\cdot)$. The authors find \edit{$K=12$} to work well for phoneme classification.

>>>>

\subsubsection{Approaches}
\paragraph{CPC}
\label{par:cpc}
Contrastive Predictive Coding (CPC)~\cite{oord_representation_2018} is a prominent example of a contrastive model. 
\edit{CPC uses a convolutional module $f_1(\cdot)$ to produce localized representations $z_t$ with a recurrent module $f_2(\cdot)$ on top that outputs a contextualized representation $h_t$. An anchor representation $\hat{z}_{t,k}$ is obtained via a linear projection $g_k(\cdot)$ of $h_t$. The positives and negatives are sampled from the localized representation $Z$.
Hence, at a single timestep $t$, CPC forms multiple anchor representations $\hat{z}_{t,k}$ for $k\in\{1,\dots,K\}$ and associates with each one a single positive sample at the corresponding timestep, $z_{t+k}$, $k$ steps in the future:
\begin{align}
    z_t &= f_{1}(X_{[t-u,t+u]}) \label{eq: cpc local representation} \enspace ,\\
    H_{[1,t]} &= f_{2}(Z_{[1,t]}) \enspace , \\
    \hat{z}_{t,k} &= g_k(h_{t}) \enspace.
\end{align}
%
\noindent Each $z_{t}$ only encodes information from a limited receptive field, while $f_2(\cdot)$ is limited to condition each $h_t$ on previous timesteps $Z_{[1,t]}$. \ci{Without these restrictions}, the model could collapse to a trivial solution. $g_k$ is a unique transformation per \ci{offset} $k$ (e.g., a linear projection).
The loss function measures the similarity between the anchor representation $\hat{z}_{t,k}$ and the positive $z_{t+k}$ normalized by the total similarity to the positive and negatives.}
The approach is similar to previous work on Noise-Contrastive Estimation (NCE)~\cite{gutmann_noisecontrastive_2010}. \edit{Minimizing} the loss corresponds to maximizing a lower bound on the mutual information between $h_t$ and $z_{t+k}$ (and in turn $x_{t+k-u:t+k+u}$) and is hence called InfoNCE:
\begin{align}
    \mathcal{L}_{t,k} &= - \log \left(\frac{\exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{t+k})}{\sum_{\ci{i \in \mathcal{I}}} \exp(\hat{z}_{t,k}^{\text{\tiny T}}z_{i})} \right)\enspace .
    \label{eq: cpc loss}
\end{align}
\noindent \edit{Here, $\mathcal{I}$ \ci{is a random subset of $N$} indices  which includes the target index $t+k$ and $N-1$ negative samples drawn from a proposal distribution, e.g., a uniform distribution over $\{1,\dots,T\}$. Including the target index in $\mathcal{I}$ ensures that the loss is a proper categorical cross-entropy and that minimizing it has the previously stated relation to mutual information maximization}. This corresponds to sampling negatives from the same sequence and has been \edit{shown} to give good performance for phoneme classification~\cite{oord_representation_2018}. The loss is indexed by $k$ to show that CPC targets multiple offsets using different projection layers $g_k(\cdot)$. The authors find \edit{$K=12$} to work well for phoneme classification.
